### 2024-04-23 parts/prelim.tex
move definitions to preliminaries

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}.

@@

\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
defined as the minimal $k$ such that $\card{f(w)} =
\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp. $\reintro*\NPoly[k]$) for
the functions in $\ZPoly$ (resp. $\NPoly$) that have growth rate at most $k$.
It is clear that a function $f \in \ZPoly$ defined using a production function
$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
generally it is known that \kl{polyregular functions} have \kl{polynomial
growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
growth}, a typical example being the function $a^n \mapsto 2^n$.

@@

\AP
As a model of computation, we will use the following definition of transducers,
that is based on the marble transducers \cite{EHB99}, and follows the general
pattern of transducers introduced in \cite{CDTL23}. Note that we only consider
$\Nat$-valued functions, but could consider other commutative semigroups.

>>>>

\AP Because the functions of interest in this paper have output in $\Nat$ or
$\Rel$, we will only provide the definition of \kl{polyregular functions} for
these two output semigroups, and we refer the reader to \cite{BOKL19} for the
general definition of \kl{polyregular functions}. As a model of computation, we
will use the following definition of transducers, that is based on the marble
transducers \cite{EHB99}, and follows the general pattern of transducers
introduced in \cite{CDTL23}. Note that we only consider $\Rel$-valued (resp.
$\Nat$-valued) functions, but could consider other commutative semigroups.

\AP It is already known that this notion of transducers exactly captures $\NSF$
and $\NPoly$, and generalizes to outputs into other commutative semigroups than
$\Nat$. To make the above statement formal, let us recall that a
\intro{counter} in an automaton is a pair $u,w$ of words together with an
integer $n > 1$, such that $\delta(q_0, uw^n) = \delta(q_0, u)$ and
$\delta(q_0, u w^k) \neq \delta(q_0, u)$ for all $k < n$; where we used the
extension of $\delta \colon Q \times \Sigma \to Q$ to words by defining
$\delta(q, \varepsilon) = q$, and $\delta(q,au) = \delta(\delta(q,a), u)$. An
automaton is \reintro{counter-free} if it contains no \reintro{counters}. This
notion can be lifted to \kl{$\mathcal{H}$-transducers} by considering counters
of their underlying automata.

\begin{lemma}[{\cite[Theorems 5.15 and 7.10]{DOUE23}}]
    \label{transducer-nsf-npoly:lemma}
    Let $f \colon \Sigma^* \to \Nat$, and let $k \in \Nat$.
    Then, 
    $f \in \NPoly[k]$ (respectively $\NSF[k]$) if and only if
    $f$ is \kl{computed} by an \kl{$\NPoly[k-1]$-transducer}
        (respectively by an \kl{$\NSF[k-1]$-transducer} that is 
        \kl{counter-free}).
\end{lemma}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

            \draw[->] (A) to node[midway, below] {$a \mid 0$} (B);

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
    \label{non-canonical-transd:fig}
\end{figure}

\begin{example}
    \label{non-canonical-transd:ex}
    Let $\BadExOk \colon \set{a}^* \to \Nat$
    be defined as
    $\BadExOk(\varepsilon) = 1$,
    $\BadExOk(a) = 0$,
    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
    depicted in \cref{non-canonical-transd:fig}.
\end{example}

\AP The \intro{growth rate} of a function $f \colon \Sigma^* \to \Rel$ is
defined as the minimal $k$ such that $\card{f(w)} =
\bigO\left(\card{w}^k\right)$. If such a $k$ exists, we say that the function
$f$ has \intro{polynomial growth}. We write $\reintro*\ZPoly[k]$ (resp.
$\reintro*\NPoly[k]$) for the functions in $\ZRat$ (resp. $\NRat$) that have
growth rate at most $k$.

It is clear that a function $f \in \ZPoly$ defined using a production function
$\pi \colon M^k \to \Rel$ has \kl{growth rate} at most $k-1$, and more
generally it is known that \kl{polyregular functions} have \kl{polynomial
growth} \cite{BOJA22}. Not all \kl{$\Nat$-rational series} have \kl{polynomial
growth}, a typical example being the function $a^n \mapsto 2^n$.

\AP
The following
\cref{nat-rel-poly:def} is one of the equivalent definitions of \cite{CDTL23},
and is similar in shape to the \emph{finite counting automata} introduced by
\cite{SCHU62}.

@@

@@
